# Exponential family of distributions

```{r setup}
#| include: false
library(ggplot2)
```

```{r}
#| echo: false
#| warning: false
# Define the x-values for the plots
x <- seq(0, 10, by = 0.1)

# Define the exponential family representations for each distribution
gaussian_pdf <- dnorm(x, mean = 5, sd = 1)
poisson_pdf <- dpois(x, lambda = 3)
binomial_pdf <- dbinom(x, size = 10, prob = 0.3)
beta_pdf <- dbeta(x, shape1 = 2, shape2 = 5)
gamma_pdf <- dgamma(x, shape = 2, scale = 2)

# Create plots for each distribution
plot_gaussian <- ggplot(data = data.frame(x, gaussian_pdf), aes(x = x, y = gaussian_pdf)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(title = "Gaussian (Normal) Distribution",
       x = "x",
       y = "Probability Density") +
  theme_minimal()

plot_poisson <- ggplot(data = data.frame(x, poisson_pdf), aes(x = x, y = poisson_pdf)) +
  geom_line(color = "green", linewidth = 1) +
  labs(title = "Poisson Distribution",
       x = "x",
       y = "Probability Mass") +
  theme_minimal()

plot_binomial <- ggplot(data = data.frame(x, binomial_pdf), aes(x = x, y = binomial_pdf)) +
  geom_line(color = "red", linewidth = 1) +
  labs(title = "Binomial Distribution",
       x = "x",
       y = "Probability Mass") +
  theme_minimal()

plot_beta <- ggplot(data = data.frame(x, beta_pdf), aes(x = x, y = beta_pdf)) +
  geom_line(color = "purple", linewidth = 1) +
  labs(title = "Beta Distribution",
       x = "x",
       y = "Probability Density") +
  theme_minimal()

plot_gamma <- ggplot(data = data.frame(x, gamma_pdf), aes(x = x, y = gamma_pdf)) +
  geom_line(color = "orange", linewidth = 1) +
  labs(title = "Gamma Distribution",
       x = "x",
       y = "Probability Density") +
  theme_minimal()
```


The exponential family is a broad class of probability distributions that includes well-known distributions such as the Gaussian (Normal), Poisson, Binomial, Beta, and Gamma distributions. These distributions can be expressed in the form of the exponential family, sharing a common structure.

## Gaussian (Normal) Distribution

The Gaussian distribution is characterized by a bell-shaped curve and is widely used in various fields of science and statistics [@GaussianCitation]. Its probability density function (PDF) is defined as:


\begin{equation}
f(x|\mu,\sigma^2) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right)
\end{equation}


## Poisson Distribution

The Poisson distribution models events occurring in fixed intervals of time or space. It is another example of an exponential family distribution when expressed in the right form.

## Binomial Distribution

The binomial distribution describes the number of successes in a fixed number of Bernoulli trials. When appropriately parameterized, it is also a member of the exponential family.

## Beta Distribution

The beta distribution is used for modeling probabilities or proportions. It, too, can be derived from the exponential family when the parameters align with the framework.

## Gamma Distribution

The gamma distribution is often used to model wait times and can be expressed as a member of the exponential family by selecting the appropriate parameters.

In the figures below, you can see how these distributions align with the exponential family framework:

```{r}
#| echo: false
#| out.width: "100%"
# Display the plots of distributions
# Arrange the plots in a grid
library(gridExtra)
grid.arrange(plot_gaussian, plot_poisson, plot_binomial, plot_beta, plot_gamma, nrow = 2)
```

These graphical representations illustrate how these common distributions fit the structure of the exponential family, allowing for a unified understanding of their relationships and facilitating statistical inference and modeling.
